---
title: "Lab Assignment 1 BIS 505b"
author: "Wenxin Xu"
date: "2/21/2021"
output: 
    html_document:
        toc: true
        highlight: default
---

<!--- Copyright (c) 2021, M. Ciarleglio --->

<!--- Set global options that apply to every code chunk in this file, can be overwritten in individual chunk headers --->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
options(scipen=999)               # global option that requests no scientific notation in this session (*optional*)
```

<!--- Load packages (previously installed)  --->
```{r, message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
library(dplyr)
```

<!--- Set WD and read data file --->
```{r, include=FALSE}
setwd("/Users/xuwenxin/Desktop/BIS 505/Assignments/Lab Assignments")    # set working directory
hgb <- read.csv("hgb.csv") 
```


# Instructions

This Lab Assignment uses the data from the study conducted to investigate the impacts of herbicide exposure on maternal health described in **Lab Assignment 0**, `hgb.csv`. The analyses that you perform will use data from the tap water consumption group (`group` = 1) and the bottled water consumption group (`group` = 2).  The women who consumed only tap water (`group` = 1) are considered to be “exposed” because they were exposed to the herbicides found in the region’s groundwater.  

When testing, use either an *unpooled t-test* (default in `t.test()`) or *continuity-corrected binomial test of proportions* (default in `prop.test()`)/*continuity-corrected chi-square test* (default in `chisq.test()`), as appropriate. Report any p-values that are less than 0.001 as <0.001 and round values reported in your narrative text to 3 decimal places. 


# Assignment

**1.** Begin with data management. [Note: No written responses are required for this question. Display the code chunks that perform the requested data management steps for this question.] 

**a.** <span style="color: #CC0000;">[5 points]</span> Import the CSV file `hgb.csv` in the third code chunk above. Name your data frame `hgb` and re-create the variables `wtgain`, `change`,`ed`, `anemic9`, `group_factor`, `parity_factor`, `prenatal_factor`, `psmoke_factor`, `ed_factor`, and `anemic9_factor` that you created in **Lab Assignment 0**. After these steps, `hgb` should contain 23 variables. 

```{r}
hgb$wtgain <- hgb$wt1 - hgb$wt0

hgb$change <- hgb$hgb36 - hgb$hgb9

hgb$ed[hgb$edyrs < 12] <- 0

hgb$ed[hgb$edyrs == 12] <- 1

hgb$ed[hgb$edyrs > 12] <- 2

hgb$anemic9[hgb$hgb9 >= 11] <- 0

hgb$anemic9[hgb$hgb9 < 11] <- 1


hgb <- dplyr::mutate(hgb,
                     group_factor=factor(group,
                                         levels=c(1,2,3),
                                         labels=c("Tap Water","Bottled/Filtered Water","Tap/Bottled/Filtered")),
                     
                     parity_factor=factor(parity,
                                          levels=c(0,1,2,3),
                                          labels=c("None","One","Two","Three or more"),
                                          ordered=TRUE),
                     
                     prenatal_factor=factor(prenatal,
                                            levels=c(1,0),
                                            labels=c("Yes","No")),
                     
                     psmoke_factor=factor(psmoke,
                                          levels=c(1,0),
                                            labels=c("Yes","No")),
                     ed_factor=factor(ed,
                                      levels=c(0,1,2),
                                      labels=c("Less than HS","HS/GHD","Some college or more"),
                                      ordered=TRUE),
                     
                     anemic9_factor=factor(anemic9,
                                           levels=c(1,0),
                                           labels=c("Anemic","Not anemic")))
```

```{r}
ncol(hgb)
```

Answer: Now `hgb` contains `r ncol(hgb)` variables.

**b.** <span style="color: #CC0000;">[5 points]</span> Create a subset called `hgb12` that only includes participants in the tap water only consumption group and the bottled/filtered water only consumption group. You will work with this data frame in this Lab Assignment. After you create `hgb12`, run the code, `hgb12 <- droplevels(hgb12)`.

```{r}
hgb12 <- subset(hgb, group == 1 | group == 2)

hgb12 <- droplevels(hgb12) # drop any unused levels from factors in the data frame
```


**2**. We would like to determine if certain characteristics differ significantly between these two water consumption groups. Our main comparison of interest is the difference in hemoglobin change during pregnancy in the tap water only group and the bottled/filtered water only group.

**a.** <span style="color: #CC0000;">[10 points]</span> Determine if pregnant women who drink only tap water and pregnant women who drink only bottled water are different, on average, with respect to the primary outcome, hemoglobin change over the course of pregnancy. **(i)** State the null and alternative hypotheses of this test and **(ii)** report the name of the test that you performed. **(iii)** From your **R** output, report the value of the test statistic and p-value. **(iv)** State your statistical conclusion and your conclusion in the context of the problem. **(v)** If a significant difference is found, state which group has a larger average hemoglobin decline and report the average decline in each group. 



```{r}
tt = t.test(change ~ group_factor, data = hgb12, var.equal=FALSE, alternative="two.sided")

tt
```

(i) $H_0: \mu_1=\mu_2$ vs. $H_1: \mu_1 \neq \mu_2$
(ii) Perform a two-sample unpooled 2-sided t-test
(iii) t statistic = `r round(tt$statistic, 3)`; p-value = `r round(tt$p.value, 3)`
(iv) Reject $H_0: \mu_1=\mu_2$ and conclude $H_1: \mu_1 \neq \mu_2$. That is, there is evidence to conclude that mean hemoglobin change is significantly different in pregnant women who drink only tap water and pregnant women who drink only bottled water (p-value from unpooled test `r round(tt$p.value, 3)` is less than significant level 0.05).
(v) Tap water consumption group has a larger average hemoglobin decline (`r round(tt$estimate[1], 3)` g/dL) than bottled/filtered water consumption group (`r round(tt$estimate[2], 3)` g/dL).


**b.** <span style="color: #CC0000;">[10 points]</span> Report the point estimate of the difference in the mean hemoglobin change between the tap water consumption group and the bottled/filtered water consumption group and the 95% confidence interval for the difference in means. Does the result of your hypothesis test in **2(a)** seem reasonable given the confidence interval for the difference in means? Why?

The point estimate of the difference in the mean hemoglobin change between the tap water consumption group and the bottled/filtered water consumption group is `r round(tt$estimate[1] - tt$estimate[2], 3)` g/dL. The 95% unpooled confidence interval for the mean difference is (`r round(tt$conf.int[1], 3)`, `r round(tt$conf.int[2], 3)`). The result of my hypothesis test in 2(a) seem reasonable given the confidence interval, because this confidence interval does not contain 0, the value hypothesized for the difference in means under $H_0$ (i.e., $H_0: \mu_1-\mu_2=0$).

**c.** Determine if mean **(i)** age, **(ii)** income, **(iii)** weight gain, **(iv)** week 9 hemoglobin, and **(v)** final hemoglobin are significantly different in the two water consumption groups. Report the name of the test performed, state your statistical conclusion, p-value, and your conclusion in the context of the problem. If a significant difference is found, state which group has a higher or lower average value of the variable being compared, report a point estimate of the difference in means, and report the 95% confidence interval for the difference in means.

  - **(i).** <span style="color: #CC0000;">[5 points]</span> Age
  
```{r}
tt_age = t.test(age ~ group_factor, data = hgb12, var.equal=FALSE, alternative="two.sided")

tt_age
```

Perform a two-sample unpooled 2-sided t-test.Reject $H_0: \mu_1=\mu_2$ and conclude $H_1: \mu_1 \neq \mu_2$. That is, there is evidence to conclude that mean age is significantly different in pregnant women who drink only tap water and pregnant women who drink only bottled water (p-value from unpooled test `r round(tt_age$p.value, 3)` is less than significant level 0.05).Tap water consumption group has a smaller average age (`r round(tt_age$estimate[1], 3)` years old) than bottled/filtered water consumption group (`r round(tt_age$estimate[2], 3)` years old). The point estimate of the difference in the mean age is `r round(tt_age$estimate[1] - tt_age$estimate[2], 3)` years old. The 95% unpooled confidence interval for the mean difference is (`r round(tt_age$conf.int[1], 3)`, `r round(tt_age$conf.int[2], 3)`).

  - **(ii).** <span style="color: #CC0000;">[5 points]</span> Income (multiply point estimate of difference in means and its and CI by 10,000 to report on the scale of dollars)

```{r}
tt_income = t.test(income ~ group_factor, data = hgb12, var.equal=FALSE, alternative="two.sided")

tt_income
```

Perform a two-sample unpooled 2-sided t-test.Reject $H_0: \mu_1=\mu_2$ and conclude $H_1: \mu_1 \neq \mu_2$. That is, there is evidence to conclude that mean income is significantly different in pregnant women who drink only tap water and pregnant women who drink only bottled water (p-value from unpooled test `r round(tt_income$p.value, 3)` is less than significant level 0.05).Tap water consumption group has a lower average income (`r round(10000*(tt_income$estimate[1]), 3)` dollars) than bottled/filtered water consumption group (`r round(10000*(tt_income$estimate[2]), 3)` dollars). The point estimate of the difference in the mean income is `r round(10000*(tt_income$estimate[1] - tt_income$estimate[2]), 3) ` dollars. The 95% unpooled confidence interval for the mean difference is (`r round(10000*(tt_income$conf.int[1]), 3)`, `r round(10000*(tt_income$conf.int[2]), 3)`).

  - **(iii).** <span style="color: #CC0000;">[5 points]</span> Weight Gain

```{r}
tt_wtgain = t.test(wtgain ~ group_factor, data = hgb12, var.equal=FALSE, alternative="two.sided")

tt_wtgain
```

Perform a two-sample unpooled 2-sided t-test.Reject $H_0: \mu_1=\mu_2$ and conclude $H_1: \mu_1 \neq \mu_2$. That is, there is evidence to conclude that mean weight gain is significantly different in pregnant women who drink only tap water and pregnant women who drink only bottled water (p-value from unpooled test `r round(tt_wtgain$p.value, 3)` is less than significant level 0.05).Tap water consumption group has a smaller average weight gain (`r round(tt_wtgain$estimate[1], 3)` lb) than bottled/filtered water consumption group (`r round(tt_wtgain$estimate[2], 3)` lb). The point estimate of the difference in the mean weight gain is `r round(tt_wtgain$estimate[1] - tt_wtgain$estimate[2], 3)` lb. The 95% unpooled confidence interval for the mean difference is (`r round(tt_wtgain$conf.int[1], 3)`, `r round(tt_wtgain$conf.int[2], 3)`).

  - **(iv).** <span style="color: #CC0000;">[5 points]</span> Week 9 Hemoglobin 

```{r}
tt_hgb9 = t.test(hgb9 ~ group_factor, data = hgb12, var.equal=FALSE, alternative="two.sided")

tt_hgb9
```


Perform a two-sample unpooled 2-sided t-test.Reject $H_0: \mu_1=\mu_2$ and conclude $H_1: \mu_1 \neq \mu_2$. That is, there is evidence to conclude that mean week 9 hemoglobin is significantly different in pregnant women who drink only tap water and pregnant women who drink only bottled water (p-value from unpooled test `r round(tt_hgb9$p.value, 3)` is less than significant level 0.05).Tap water consumption group has a smaller average week 9 hemoglobin (`r round(tt_hgb9$estimate[1], 3)` g/dL) than bottled/filtered water consumption group (`r round(tt_hgb9$estimate[2], 3)` g/dL). The point estimate of the difference in the mean week 9 hemoglobin is `r round(tt_hgb9$estimate[1] - tt_hgb9$estimate[2], 3)` g/dL. The 95% unpooled confidence interval for the mean difference is (`r round(tt_hgb9$conf.int[1], 3)`, `r round(tt_hgb9$conf.int[2], 3)`).


  - **(v).** <span style="color: #CC0000;">[5 points]</span> Final Hemoglobin 

```{r}
tt_hgb36 = t.test(hgb36 ~ group_factor, data = hgb12, var.equal=FALSE, alternative="two.sided")

tt_hgb36
```

Perform a two-sample unpooled 2-sided t-test.Reject $H_0: \mu_1=\mu_2$ and conclude $H_1: \mu_1 \neq \mu_2$. That is, there is evidence to conclude that mean final hemoglobin is significantly different in pregnant women who drink only tap water and pregnant women who drink only bottled water (p-value from unpooled test `r round(tt_hgb36$p.value, 3)` is less than significant level 0.05).Tap water consumption group has a smaller average final hemoglobin (`r round(tt_hgb36$estimate[1], 3)` g/dL) than bottled/filtered water consumption group (`r round(tt_hgb36$estimate[2], 3)` g/dL). The point estimate of the difference in the mean final hemoglobin is `r round(tt_hgb36$estimate[1] - tt_hgb36$estimate[2], 3)` g/dL. The 95% unpooled confidence interval for the mean difference is (`r round(tt_hgb36$conf.int[1], 3)`, `r round(tt_hgb36$conf.int[2], 3)`).


**Total answer**: Mean age, income, weight gain, week 9 hemoglobin and final hemoglobin are all significantly different in the two water consumption groups.


**3**. Another way of looking at hemoglobin is to dichotomize into an indicator of presence of anemia. You created this indicator in question **1(a)** using the week 9 hemoglobin value `anemic9_factor`. 

**a.** <span style="color: #CC0000;">[10 points]</span> Determine if the prevalence of anemia ("disease") at week 9 of pregnancy is significantly different in pregnant women who drink only tap water ("exposed") and pregnant women who drink only bottled/filtered water ("unexposed"). **(i)** State the null and alternative hypotheses of this test and **(ii)** report the name of the test that you performed. **(iii)** From your **R** output, report the value of the test statistic and p-value. **(iv)** State your statistical conclusion and your conclusion in the context of the problem. **(v)** If a significant difference is found, state which group has a larger estimated prevalence of anemia at week 9 and report the proportion with anemia at week 9 in each group. 


```{r}
tab2p = table(hgb12$group_factor, hgb12$anemic9_factor,
              dnn = c("Water consumption", "Anemic"))
tab2p

```


```{r}
pt = prop.test(tab2p,alternative = "two.sided") 

pt
```


(i) $H_0: p_1=p_2$ vs. $H_1: p_1 \neq p_2$
(ii) Perform a two-sample binomial test of proportions
(iii) Chi statistic = `r round(pt$statistic, 3)`; p-value = `r round(pt$p.value, 3)`
(iv) Reject $H_0: p_1=p_2$ and conclude $H_1: p_1 \neq p_2$. That is, there is evidence to conclude that the prevalence of anemia at week 9 of pregnancy is significantly different in pregnant women who drink only tap water ("exposed") and pregnant women who drink only bottled/filtered water ("unexposed") (p-value `r round(pt$p.value, 3)` is less than significant level 0.05).
(v) Tap water consumption group has a higher estimated prevalence of anemia at week 9 (`r round(pt$estimate[1], 3)`) than bottled/filtered water consumption group (`r round(pt$estimate[2], 3)`).


**b.** <span style="color: #CC0000;">[10 points]</span> Report the point estimate of the difference in the proportion of women with anemia at week 9 of pregnancy between the tap water consumption group and the bottled/filtered water consumption group (a.k.a., the risk difference or attributable risk). Report the 95% confidence interval for the difference in proportions.


The point estimate of the difference in the proportion of women with anemia at week 9 of pregnancy between the tap water consumption group and the bottled/filtered water consumption group is `r round(pt$estimate[1] - pt$estimate[2], 3)`. The 95% confidence interval for the mean difference is (`r round(pt$conf.int[1], 3)`, `r round(pt$conf.int[2], 3)`).


**c.** Determine there is an association between tap or bottled/filtered water consumption and **(i)** receiving adequate prenatal care, **(ii)** smoking status prior to pregnancy, **(iii)** educational attainment, and **(iv)** parity using a chi-square test. State your statistical conclusion, p-value, and your conclusion in the context of the problem. 

  - **(i).** <span style="color: #CC0000;">[5 points]</span> Adequate prenatal care
  
```{r}
tab2p_prenatal = table(hgb12$group_factor, hgb12$prenatal_factor,
              dnn = c("Water consumption", "Adequate prenatal care"))

tab2p_prenatal
```

```{r}
ct_prenatal = chisq.test(tab2p_prenatal)

ct_prenatal
```

Reject $H_0$ and conclude $H_1$. That is, there is a significant association between tap or bottled/filtered water consumption and receiving adequate prenatal care (p-value = `r round(ct_prenatal$p.value, 3)`). 

  - **(ii).** <span style="color: #CC0000;">[5 points]</span> Smoking status

```{r}
tab2p_psmoke = table(hgb12$group_factor, hgb12$psmoke_factor,
              dnn = c("Water consumption", "Smoking status"))

tab2p_psmoke
```

```{r}
ct_psmoke = chisq.test(tab2p_psmoke)

ct_psmoke
```

Reject $H_0$ and conclude $H_1$. That is, there is a significant association between tap or bottled/filtered water consumption and smoking status prior to pregnancy (p-value = `r round(ct_psmoke$p.value, 3)`). 

  - **(iii).** <span style="color: #CC0000;">[5 points]</span> Educational attainment

```{r}
tab2p_ed = table(hgb12$group_factor, hgb12$ed_factor,
              dnn = c("Water consumption", "Educational attainment"))

tab2p_ed
```

```{r}
ct_ed = chisq.test(tab2p_ed)

ct_ed
```

Reject $H_0$ and conclude $H_1$. That is, there is a significant association between tap or bottled/filtered water consumption and educational attainment (p-value = `r round(ct_ed$p.value, 3)`). 

  - **(iv).** <span style="color: #CC0000;">[5 points]</span> Parity

```{r}
tab2p_parity = table(hgb12$group_factor, hgb12$parity_factor,
              dnn = c("Water consumption", "Parity"))

tab2p_parity
```

```{r}
ct_parity = chisq.test(tab2p_parity)

ct_parity
```

Fail to reject $H_0$. That is, there is not a significant association between tap or bottled/filtered water consumption and parity (p-value = `r round(ct_parity$p.value, 3)`). 

  
**4.** <span style="color: #CC0000;">[5 points]</span> We have identified several significant differences between the two water consumption groups of interest. In particular, we found that one group tends to have a larger average decrease in hemoglobin. Since we would like to attribute this difference to herbicide exposure through drinking water, are you concerned by any of the differing demographic/baseline characteristics between the two groups?  That is, do you think differences in certain baseline characteristics will make it more difficult to attribute the differences in hemoglobin change to herbicide exposure alone?  

Yes. We found that mean age, income, weight gain, week 9 hemoglobin and final hemoglobin are all significantly different in the two water consumption group from 2(c), and adequate prenatal care, smoking status and educational attainment all have significant association with water consumption. Therefore, those 8 baseline characteristics will make it more difficult to attribute the differences in hemoglobin change to herbicide exposure alone.

