---
title: "Lab Assignment 4 BIS 505b"
author: "Wenxin Xu"
date: "3/28/2021"
output:
  pdf_document:
    toc: yes
  html_document:
    highlight: default
    toc: yes
---

<!--- Copyright (c) 2021, M. Ciarleglio --->

<!--- Set global options that apply to every code chunk in this file, can be overwritten in individual chunk headers --->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
options(show.signif.stars=FALSE)  # global option to hide significance stars (*optional*)
```

<!--- Load packages (previously installed)  --->
```{r, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(car)
library(multcomp)
```

<!--- Set WD and read data file --->
```{r, include=FALSE}
#setwd(" ")    # set working directory
setwd("/Users/xuwenxin/Desktop/BIS 505/Assignments/Lab Assignments")
#hgb <-        # read data file
hgb<-read.csv("hgb.csv") 
```


# Instructions

This Lab Assignment uses the data from the study conducted to investigate the impacts of herbicide exposure on maternal health described in **Lab Assignment 0**, `hgb.csv`. We would like to explore the association between tap water consumption group and hemoglobin change (g/dL), controlling for potentially important variables or confounders of the association using multiple linear regression. In this assignment, report any p-values that are less than 0.001 as **<0.001** and round values reported in your narrative text to **3** decimal places. **Be sure to clearly state the reference category when interpreting the effects of categorical variables in any regression model.**

# Assignment

**1.** <span style="color: #CC0000;">[5 points]</span> Import the CSV file `hgb.csv` in the third code chunk above. Name your data frame `hgb` and re-create the variables `group_factor` (reference = Bottled only (code provided below)), `prenatal_factor` (reference = No) and `psmoke_factor` (reference = No) that you created in **Lab Assignment 0**.  

Instead of analyzing `change` in this lab, we will analyze `hgbdecline = hgb$hgb9 - hgb$hgb36`, or week 9 hemoglobin [`hgb9`] minus the week 36 hemoglobin [`hgb36`]. Since hemoglobin decreases during pregnancy for all women in our data set, the variable `hgbdecline` will be positive for all individuals. A larger value for `hgbdecline` indicates that hemoglobin decreased a greater amount during pregnancy. When modeling `hgbdecline` as the response, positive slopes indicate that hemoglobin is declining more (greater decline from baseline), while negative slopes indicate that hemoglobin is declining less (smaller decline from baseline). 

After these steps, `hgb` should contain 17 variables. [**Note:** When creating factor variables, **do not** use the `ordered=TRUE` option to create ordinal variables. No written response is required for this question. Display the code chunk(s) that perform the requested data management steps for this question.] 
```{r}
hgb$hgbdecline <- hgb$hgb9 - hgb$hgb36

hgb$group_factor <- factor(hgb$group, 
                           levels = c(2, 3, 1),
                           labels = c("Bottled only",
                                      "Combination",
                                      "Tap only"))
hgb$prenatal_factor <- factor(hgb$prenatal,
                              levels = c(0,1),
                              labels=c("No","Yes"))

hgb$psmoke_factor <- factor(hgb$psmoke,
                            levels=c(0,1),
                           labels = c("No","Yes"))

# now data frame hgb has 17 variables
ncol(hgb)
```



**2.** The **research question** is: Is type of water consumed [`group_factor`] associated with hemoglobin decline during pregnancy [`hgbdecline`]?

**a.** <span style="color: #CC0000;">[20 points]</span> Use water consumption group [`group_factor`] to model hemoglobin decline [`hgbdecline`] using a linear regression model. 

- How many dummy variables represent `group_factor` in this model? What is the reference category of `group_factor`?
```{r}
contrasts(hgb$group_factor)
```

`group_factor` has `r ncol(contrasts(hgb$group_factor))` dummy variables. The reference category is `r rownames(contrasts(hgb$group_factor))[1]` group.


- Write the fitted model and interpret the regression parameters (intercept and two slopes). Do the slopes indicate that hemoglobin is declining more (greater decline from baseline) or less (smaller decline from baseline) in the groups being compared?

```{r}
mod.group <- lm(hgbdecline ~ group_factor, data=hgb)
summary(mod.group)
```

```{r}
confint(mod.group)
```


The fitted model is: $\hat{y}=$ `r round(summary(mod.group)$coefficients[1],3)` `r ifelse(summary(mod.group)$coefficients[2]<0,"$-$","$+$")` `r round(abs(summary(mod.group)$coefficients[2]),3)` Combination `r ifelse(summary(mod.group)$coefficients[3]<0,"$-$","$+$")` `r round(abs(summary(mod.group)$coefficients[3]),3)` Tap only.

The average hemoglobin decline in Bottled only group is $a=$ `r round(summary(mod.group)$coefficients[1],3)` [95% CI (`r round(confint(mod.group)[1,1],3)`,`r round(confint(mod.group)[1,2],3)`)]

The average hemoglobin decline in Combination group is $b_1=$ `r round(summary(mod.group)$coefficients[2],3)` [95% CI (`r round(confint(mod.group)[2,1],3)`, `r round(confint(mod.group)[2,2],3)`)] units higher than that in Bottled only group.

The average hemoglobin decline in Tap only group is $b_2=$ `r round(summary(mod.group)$coefficients[3],3)` [95% CI (`r round(confint(mod.group)[3,1],3)`, `r round(confint(mod.group)[3,2],3)`)] units higher than that in Bottled only group. 

- Perform three hypothesis tests each at the $\alpha$ = 0.05-level to test the following null hypotheses: (1) $H_0: \beta_1 = 0$, (2) $H_0: \beta_2 = 0$, and (3) $H_0: \beta_1 = \beta_2 = 0$. For each test, (i) State the null and alternative hypotheses; (ii) From your **R** output, report the value of the test statistic and p-value; (iii) State your statistical conclusion and your conclusion in the context of the problem.

(1) $H_0: \beta_1 = 0$ vs. $H_1: \beta_1 \neq 0$

T test statistic is `r round(summary(mod.group)$coefficients[8], 3)`, p value is `r ifelse(summary(mod.group)$coefficients[11]<.001, "<.001",paste("=", round(summary(mod.group)$coefficients[11],3)))`. We reject $H_0$ and conclude that there is a significant difference in the average hemoglobin decline in Combination group vs. Bottled only group.

(2) $H_0: \beta_2 = 0$ vs. $H_1: \beta_2 \neq 0$

T test statistic is `r round(summary(mod.group)$coefficients[9], 3)`, p-value `r ifelse(summary(mod.group)$coefficients[12]<.001, "<.001",paste("=", round(summary(mod.group)$coefficients[12],3)))`. We reject $H_0$ and conclude that there is a significant difference in the average hemoglobin decline in Combination group vs. Bottled only group.


(3) $H_0: \beta_1 = \beta_2 = 0$ vs. $H_1: \beta_1, \beta_2$ not all 0.

The F-statistic is `r round(summary(mod.group)$fstatistic[1],3)`, p-value `r ifelse(as.numeric(1-pf(summary(mod.group)$fstatistic[1], summary(mod.group)$fstatistic[2], summary(mod.group)$fstatistic[3]))<.001,"<.001",paste("=",round(as.numeric(1-pf(summary(mod.group)$fstatistic[1], summary(mod.group)$fstatistic[2], summary(mod.group)$fstatistic[3])),3)))`. We reject $H_0$ and conclude that there is at least one significant difference in the average hemoglobin decline in Combination group vs. Bottled only group or Tap only group vs. Bottled only group.



**b.** <span style="color: #CC0000;">[30 points]</span> Perhaps the effect we observe due to water consumption group is driven by demographic characteristics. For example, maybe women who drink only tap water were more likely to be pre-pregnancy smokers or are less likely to receive adequate prenatal care, and these factors are driving the larger decline in hemoglobin in this group. Our goal is to control for additional characteristics of the mother and examine the *adjusted* effect of type of water consumed [`group_factor`]. Build a multiple linear regression model that controls for income [`income`], number of previous births [`parity`], adequate prenatal care [`prenatal_factor`], and pre-pregnancy smoking status [`psmoke_factor`]. 


```{r}
mod.mlr <- lm(hgbdecline ~ group_factor + income + parity + prenatal_factor + psmoke_factor, data= hgb)
summary(mod.mlr)
```

```{r}
confint(mod.mlr)
```

- Write the fitted model and interpret the regression parameters (slopes). Based on the p-values reported in the **R** output for each slope, state your conclusion about each effect in the MLR model in the context of the problem.

The fitted model is $\hat y=$ `r round(summary(mod.mlr)$coefficients[1],3)` `r ifelse(summary(mod.mlr)$coefficients[2]<0, "$-$","$+$")` `r round(abs(summary(mod.mlr)$coefficients[2]),3)` Combination `r ifelse(summary(mod.mlr)$coefficients[3]<0, "$-$","$+$")` `r round(abs(summary(mod.mlr)$coefficients[3]),3)` Tap only `r ifelse(summary(mod.mlr)$coefficients[4]<0, "$-$","$+$")` `r round(abs(summary(mod.mlr)$coefficients[4]),3)` Income `r ifelse(summary(mod.mlr)$coefficients[5]<0, "$-$","$+$")` `r round(abs(summary(mod.mlr)$coefficients[5]),3)` Parity `r ifelse(summary(mod.mlr)$coefficients[6]<0, "$-$","$+$")` `r round(abs(summary(mod.mlr)$coefficients[6]),3)` Prenatal `r ifelse(summary(mod.mlr)$coefficients[7]<0, "$-$","$+$")` `r round(abs(summary(mod.mlr)$coefficients[7]),3)` Psmoke.

**Income, number of previous births, adequate prenatal care and pre-pregnancy smoking status-adjusted effect of water consumption on hemoglobin decline:**

The estimated slope of $b_1$ indicates that the average hemoglobin decline in Combination group is `r round(summary(mod.mlr)$coefficients[2],3)` [95% CI (`r round(confint(mod.mlr)[2,1],3)`, `r round(confint(mod.mlr)[2,2],3)`)] units higher than that in Bottled only group, controlling for income, number of previous births, adequate prenatal care and pre-pregnancy smoking status.

A **significance test** of $\beta_1$ shows that there is a significant difference in the average hemoglobin decline in Combination group vs. Bottled only group (p-value `r ifelse(summary(mod.mlr)$coefficients[23]<.001,"<.001",paste("=",round(summary(mod.mlr)$coefficients[23],3)))`) when controlling for income, number of previous births, adequate prenatal care and pre-pregnancy smoking status.  

The estimated slope of $b_2$ indicates that the average hemoglobin decline in Tap only group is `r round(summary(mod.mlr)$coefficients[3],3)` [95% CI (`r round(confint(mod.mlr)[3,1],3)`, `r round(confint(mod.mlr)[3,2],3)`)] units higher than that in Bottled only group, controlling for income, number of previous births, adequate prenatal care and pre-pregnancy smoking status.

A **significance test** of $\beta_2$ shows that there is a significant difference in the average hemoglobin decline in Tap only group vs. Bottled only group (p-value `r ifelse(summary(mod.mlr)$coefficients[24]<.001,"<.001",paste("=",round(summary(mod.mlr)$coefficients[24],3)))`) when controlling for income, number of previous births, adequate prenatal care and pre-pregnancy smoking status.


- Comment on how the adjusted effects of `group_factor` on `hgbdecline` have changed in this model (adjusted model) compared to the model you constructed in question **2a** (unadjusted model). 

The adjusted effects of `group_factor` on `hgbdecline` are smaller than the unadjusted effects. $b_1$: `r round(summary(mod.mlr)$coefficients[2],3)` (adjusted) vs.`r round(summary(mod.group)$coefficients[2],3)` (unadjusted), $b_2$: `r round(summary(mod.mlr)$coefficients[3],3)` (adjusted) vs.`r round(summary(mod.group)$coefficients[3],3)` (unadjusted).


- Report the adjusted $R^2$ of this model to 4 decimal places.

The adjusted $R^2$ is `r round(summary(mod.mlr)$adj.r.squared,4)`.

**c.** <span style="color: #CC0000;">[10 points]</span> Modify the adjusted model from question **2b** to additionally control for baseline hemoglobin `hgb9`. 

```{r}
mod.mlr2 <- lm(hgbdecline ~ group_factor + income + parity + prenatal_factor + psmoke_factor+hgb9, data= hgb)
summary(mod.mlr2)
```

- Is the effect of baseline hemoglobin statistically significant in the presence of all of these other predictors? 

(i) State the null and alternative hypotheses; 
$H_0: \beta_7 = 0$ vs. $H_1: \beta_7 \neq 0$

(ii) From your **R** output, report the value of the test statistic and p-value.

The t test statistic is `r round(summary(mod.mlr2)$coefficients[24],3)`, the p-value `r ifelse(summary(mod.mlr2)$coefficients[32]<.001, "<.001", paste("=",round(summary(mod.mlr2)$coefficients[32],3)))`.

(iii) State your statistical conclusion and your conclusion in the context of the problem. 

We fail to reject $H_0$ and conclude that there is no significant association between baseline hemoglobin and average hemoglobin decline, controlling for water consumption, income, number of previous births, adequate prenatal care and pre-pregnancy smoking status.

- Report the adjusted $R^2$ of this model to 4 decimal places. Has the adjusted $R^2$ improved compared to the model in question **2b**? Is `hgb9` adding to the predictive ability of the model?

The adjusted $R^2$ is `r round(summary(mod.mlr2)$adj.r.squared,4)`. The adjusted $R^2$ decreased, so it did not improved compared to the model in question **2b**. Thus, `hgb9` is not adding to the predictive ability of the model.

**d.** <span style="color: #CC0000;">[35 points]</span> Expand on the model fit in question **2a** to include the main effect of income and an interaction term between water consumption group and income. 

```{r}
# interaction model of water consumption and income
mod.intx <- lm(hgbdecline ~ income + group_factor + income*group_factor,
               data=hgb)
summary(mod.intx)
```


- Write the fitted model. 

The fitted model is $\hat y=$ `r round(summary(mod.intx)$coefficients[1],3)` `r ifelse(summary(mod.intx)$coefficients[2]<0,"$-$","$+$")` `r round(abs(summary(mod.intx)$coefficients[2]),3)` Income `r ifelse(summary(mod.intx)$coefficients[3]<0,"$-$","$+$")` `r round(abs(summary(mod.intx)$coefficients[3]),3)` Combination `r ifelse(summary(mod.intx)$coefficients[4]<0,"$-$","$+$")` `r round(abs(summary(mod.intx)$coefficients[4]),3)` Tap only  `r ifelse(summary(mod.intx)$coefficients[5]<0,"$-$","$+$")` `r round(abs(summary(mod.intx)$coefficients[5]),3)` Income $\times$ Combination `r ifelse(summary(mod.intx)$coefficients[6]<0,"$-$","$+$")` `r round(abs(summary(mod.intx)$coefficients[6]),3)` Income $\times$ Tap only.


- Write the model $\mu_{y|x}$ in each of the three water consumption groups (bottled only group, combination group, and tap only group).

Model for bottled only group:

$\mu_{y|x_1,z_1=0,z_2=0}=\alpha+\beta_1x_1+\beta_2(0)+\beta_3(0)+\beta_4x_1(0)+\beta_5x_1(0)=\alpha+\beta_1x_1$

Model for combination group:

$\mu_{y|x_1,z_1=1,z_2=0}=\alpha+\beta_1x_1+\beta_2+\beta_3(0)+\beta_4x_1+\beta_5x_1(0)=\alpha+\beta_2+(\beta_1+\beta_4)x_1$

Model for tap only group:

$\mu_{y|x_1,z_1=0,z_2=1}=\alpha+\beta_1x_1+\beta_2(0)+\beta_3+\beta_4x_1(0)+\beta_5x_1=\alpha+\beta_3+(\beta_1+\beta_5)x_1$

- Perform a partial F-test to simultaneously test the model parameters ($\beta$s) involved in the interaction. (i) State the null and alternative hypotheses; (ii) From your **R** output, report the value of the test statistic and p-value; (iii) State your statistical conclusion and your conclusion in the context of the problem.  

```{r}
# full model is the previous interaction model

# reduced model, under H_0, don't include interaction term
mod.red <- lm(hgbdecline ~ income + group_factor,
               data=mod.intx$model)

# F-test comparing full and reduced models
anova(mod.red, mod.intx)
```


(i) State the null and alternative hypotheses; 

$H_0:\beta_4 = \beta_5 = 0$ vs. $H_1:\beta_4,\beta_5$ not all 0.

(ii) From your **R** output, report the value of the test statistic and p-value; 

The F test statistic is `r round(anova(mod.red, mod.intx)$F[2],3)`, the p-value `r ifelse(anova(mod.red, mod.intx)$"Pr(>F)"[2]<.001, "<.001", paste("=",round(anova(mod.red, mod.intx)$"Pr(>F)"[2],3)))`.

(iii) State your statistical conclusion and your conclusion in the context of the problem.  

We reject $H_0$ and conclude that there is at least one significant difference in the effect of income on hemoglobin decline in the combination group vs. bottled only group or in the tap only group vs. bottled only group.

- Tease apart the interaction to report and interpret the effect of income in each water consumption group (bottled only group, combination group, and tap only group). Perform a hypothesis test to determine if the effect of income is statistically significant in each group.



1)The effect of income on hemoglobin decline in the bottled only group is estimated by $b_1$

```{r}
# b1: effect of income in the bottled only group

# the same interaction model used in question d
mod.intx <- lm(hgbdecline ~ income + group_factor + income*group_factor,
               data=hgb)
summary(mod.intx)
```

(i) State the null and alternative hypotheses; 

$H_0:\beta_1=0$ vs. $H_1:\beta_1 \neq 0$

(ii) From your **R** output, report the p-value of this test; 

The p-value `r ifelse(summary(mod.intx)$coefficients[20]<.001, "<.001", paste("=",round(summary(mod.intx)$coefficients[20],3)))`

(iii) State your statistical conclusion and your conclusion in the context of the problem. 

We fail to reject $H_0$ and conclude that there is no significant association between income and hemoglobin decline in the bottled only group.


2) The effect of income on hemoglobin decline in the combination group is estimated by $b_1+b_4$.

```{r}
# b1 + b4: effect of income in the combination group

# vector that specifies linear combination of coefficients interested in
K1 <- rbind(c(0,1,0,0,1,0))

# label for comparison (printed in the output)
rownames(K1) <- "b1+b4 (slope in group_factor=combination)"

# estimate of slope (b1+b4) and hypothesis test
summary(glht(mod.intx,linfct=K1))
```

```{r}
# confidence interval for beta1+beta3
confint(glht(mod.intx,linfct=K1))
```


(i) State the null and alternative hypotheses; 

$H_0:\beta_1+\beta_4=0$ vs. $H_1:\beta_1+\beta_4 \neq 0$

(ii) From your **R** output, report the p-value of this test; 

The p-value `r ifelse(summary(glht(mod.intx, linfct=K1))$test$pvalues<.001,"<.001",paste("=",round(summary(glht(mod.intx, linfct=K1))$test$pvalues,3)))`. 

(iii) State your statistical conclusion and your conclusion in the context of the problem. 

We fail to reject $H_0$ and conclude that there is no significant association between income and hemoglobin decline in the combination group.


3) The effect of income on hemoglobin decline in the tap only group is estimated by $b_1+b_5$.


```{r}
# b1 + b5: effect of income in the tap only group

# vector that specifies linear combination of coefficients interested in
K2 <- rbind(c(0,1,0,0,0,1))

# label for comparison (printed in the output)
rownames(K2) <- "b1+b4 (slope in group_factor=tap only)"

# estimate of slope (b1+b5) and hypothesis test
summary(glht(mod.intx,linfct=K2))
```

```{r}
confint(glht(mod.intx,linfct=K2))
```

(i) State the null and alternative hypotheses; 

$H_0:\beta_1+\beta_5=0$ vs. $H_1:\beta_1+\beta_5 \neq 0$

(ii) From your **R** output, report the p-value of this test; 

The p-value `r ifelse(summary(glht(mod.intx, linfct=K2))$test$pvalues<.001,"<.001",paste("=",round(summary(glht(mod.intx, linfct=K2))$test$pvalues,3)))`. 

(iii) State your statistical conclusion and your conclusion in the context of the problem. 

We reject $H_0$ and conclude that there is significant association between income and hemoglobin decline in the tap only group.



 